---
title: "Introduction to taxodist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to taxodist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(taxodist)
```

## Overview

`taxodist` computes phylogenetic distances between any two taxa using
hierarchical lineage data from [The Taxonomicon](http://taxonomicon.taxonomy.nl),
a comprehensive curated classification of all life based on *Systema Naturae 2000*
(Brands, 1989 onwards).

The default distance metric is the Jaccard-based taxonomic distance:

$$d_{jaccard}(A, B) = 1 - \frac{\text{depth}(\text{MRCA}(A,B))}{\text{depth}(A) + \text{depth}(B) - \text{depth}(\text{MRCA}(A,B))}$$

This returns a value between 0 (identical lineages) and 1 (no shared ancestry
beyond root), normalized for lineage depth so that deeply resolved taxa such as
mammals are not penalized relative to shallowly resolved ones. Three methods
are available via the `method` argument: `"jaccard"` (default), `"norm"`, and
`"raw"`.

---

## Basic usage

### Getting a lineage

```{r lineage}
lin <- get_lineage("Tyrannosaurus")
print(lin)
```

### Computing distance between two taxa

```{r distance}
result <- taxo_distance("Tyrannosaurus", "Velociraptor")
print(result)
```

### Finding the most recent common ancestor

```{r mrca}
mrca("Tyrannosaurus", "Triceratops")   # Dinosauria
mrca("Tyrannosaurus", "Homo")          # Amniota
mrca("Tyrannosaurus", "Quercus")       # discaria
```

---

## Working with multiple taxa

### Pairwise distance matrix

```{r matrix}
theropods <- c("Tyrannosaurus", "Velociraptor", "Spinosaurus",
               "Allosaurus", "Carnotaurus")

mat <- distance_matrix(theropods)
print(mat)
```

### Finding the closest relative

```{r closest}
closest_relative(
  "Tyrannosaurus",
  c("Velociraptor", "Triceratops", "Brachiosaurus", "Allosaurus")
)
```

---

## Lineage utilities

### Comparing lineages side by side

```{r compare}
compare_lineages("Tyrannosaurus", "Triceratops")
```

### Listing shared clades

```{r shared}
shared_clades("Tyrannosaurus", "Velociraptor")
```

### Testing clade membership

```{r membership}
is_member("Tyrannosaurus", "Theropoda")    # TRUE
is_member("Triceratops",   "Theropoda")    # FALSE
is_member("Homo",          "Amniota")      # TRUE
```

### Filtering a list of taxa by clade

```{r filter}
taxa <- c("Tyrannosaurus", "Triceratops", "Velociraptor",
          "Brachiosaurus", "Homo", "Quercus")

filter_clade(taxa, "Dinosauria")
filter_clade(taxa, "Theropoda")
```

---

## Coverage and caching

### Checking coverage

```{r coverage}
check_coverage(c("Tyrannosaurus", "Velociraptor", "Fakeosaurus"))
```

### Caching

Lineages are automatically cached in memory during an R session to avoid
redundant network requests. Clear the cache with:

```{r cache}
clear_cache()
```

---

## Data source and citation

All lineage data is sourced from **The Taxonomicon** (taxonomy.nl), based on
*Systema Naturae 2000*:

> Brands, S.J. (1989 onwards). *Systema Naturae 2000*. Amsterdam,
> The Netherlands. Retrieved from The Taxonomicon,
> http://taxonomicon.taxonomy.nl.

Please cite this resource in any published work using `taxodist`.
